Tracing the Meta-Level: PyPy's JIT Compiler

### 2009-04-01 talk/icooolps2009/paper.tex
Small typos.

A tracing JIT compiler finds the hot loops of the program it is compiling. In
our case, this program is the language interpreter. The hot loop of the language
interpreter is its bytecode dispatch loop. Usually that is is also the only hot
loop of the language interpreter.  Tracing one iteration of this loop means that
the recorded trace corresponds to execution of one opcode. This means that the
assumption that the tracing JIT makes -- that several iterations of a hot loop
take the same or similar code paths -- is just wrong in this case. It is very
unlikely that the same particular opcode is executed many times in a row.

@@

Applying a trace-based optimizer to an interpreter, adding hints to help the
tracer produce better results been tried before in the context of the DynamoRIO
project \cite{sullivan_dynamic_2003}. This work is conceptually very close to
ours. They achieve the same unrolling of the interpreter loop so that the
unrolled version corresponds to the loops in the user program. However the
approach is greatly hindered by the fact that they trace on the machine code
level and thus have no high-level information available about the interpreter.
This makes it necessary to add quite a large number of hints, because at the
assembler level it is not really visible anymore that e.g. a bytecode string is
really immutable. Also more advanced optimizations like allocation removal would
not be possible with that approach.

>>>>

A tracing JIT compiler finds the hot loops of the program it is compiling. In
our case, this program is the language interpreter. The hot loop of the language
interpreter is its bytecode dispatch loop. Usually that is is also the only hot
loop of the language interpreter \arigo{Uh?}.  Tracing one iteration of this loop means that
the recorded trace corresponds to execution of one opcode. This means that the
assumption that the tracing JIT makes -- that several iterations of a hot loop
take the same or similar code paths -- is just wrong in this case. It is very
unlikely that the same particular opcode is executed many times in a row.

@@

Applying a trace-based optimizer to an interpreter and adding hints to help the
tracer produce better results has been tried before in the context of the DynamoRIO
project \cite{sullivan_dynamic_2003}. This work is conceptually very close to
ours. They achieve the same unrolling of the interpreter loop so that the
unrolled version corresponds to the loops in the user program. However the
approach is greatly hindered by the fact that they trace on the machine code
level and thus have no high-level information available about the interpreter.
This makes it necessary to add quite a large number of hints, because at the
assembler level it is not really visible anymore that e.g. a bytecode string is
really immutable. Also more advanced optimizations like allocation removal would
not be possible with that approach.
